Static Analysis by Path Focusing

### 2011-05-23 report/report.tex
applied David's patches

\usepackage{mathcomp,amssymb,verimagstudent}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\Q{\mathbb{Q}}
\def\P{\mathcal{P}}
\def\ite{\textnormal{ite\ }}
\def\lfp{\textnormal{\it lfp}}

@@

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discover properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

Static analysis is aimed at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
the proof of satefy properties in critical systems, such as avionics.

@@

Satisfiability Modulo Theory (SMT) solving is a field in computer science that
is aimed to decide the satisfiability of a logic formula containing Boolean
predicates and elements of a theory, such as linear arithmetic relations between
integers.
In this report, we present a novel refinement of the Abstract Interpretation
technique, that uses SMT-solving to guide the fixpoint iterations, so that it
temporarily focuses on a selected path in the control flow graph (the graph
expressing the semantics of the program), in order to
obtain more precise results at reasonable cost.

\subsection*{Contribution and Organization}

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

The report is organized as follows:
Part \ref{stateoftheartpart} introduces the state of the art in Abstract
Interpretation and more specifically in Linear Relation Analysis.
Part \ref{pathfocusingpart} describes the two techniques
\emph{Lookahead Widening} and \emph{Path Focusing}.
Part \ref{implementationpart} presents the implementation of these two
techniques into an analyzer, and the experimental results we obtained.
Finally, Part \ref{future} introduces some directions to explore.

@@

\section{Introductive example}

@@

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}

We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more elements in the sets\dots
\end{itemize}

@@

Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method for
finding approximate solutions of
fixpoint equations. This method is used for program analysis, since analyzing a
program often comes down to solving fixpoint equations, because of the loops and
recursive procedures.

@@

Abstract Interpretation method proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values,
by choosing an simpler domain  $A$ called \emph{abstract domain}.

@@

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique is aimed at computing
one of these $x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$

where $\perp$ is the least element of $A$.
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, or of finite depth, meaning that there doesn't exist
any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, in case of a domain of infinite
depth, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

Most of the domains that are used for program analysis doesn't satisfy this
finiteness property. To ensure convergence in this case, another approximation
is performed: a new operator is defined, called \emph{widening operator}, that
extrapolates the limit of a sequence of abstract values
\cite{CC77,CousotCousot92-4}. This \emph{widening
operator} is usually noted $\nabla: A \times A \rightarrow A$, and satisfies
the following properties:

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \nabla x_2$ and $x_2 \leq_A x_1
\nabla x_2$. This guarantees the correctness of the result.
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \nabla\ x_{i+1},\ \  \forall i > 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \nabla\ \Phi^\#(x'_i),\ \ \forall i > 0
}$$

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. 
Intuitively, the polyhedron $P \nabla Q$ is obtained after 
removing from the system of linear equations defining $P$ all the inequalities
that are not satisfied by $Q$.
For instance, if $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$ and $Q = 
\{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$, then the result of the widening operator
will be $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.

There have been various
work proposing a definition or a refinment of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying windening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. Most of the
cases, $n$ can be choosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\nabla_T$ is defined as follows:
$P \nabla_T Q$ is the intersection of $P \nabla Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

For instance, consider the program:
\begin{C}
while (x <= 50) {
	x++;
	y++;
}
\end{C}
If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
widening (which gives the result $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
a widening with the threshold $x \leq 50$ gives 
$P \nabla_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
with threshold helps to find better invariants.
\end{itemize}

@@

$$\left[ \begin{array}{c}
\vdots \\
X_p \\
\vdots
\end{array} \right] \longleftarrow \left[ \begin{array}{c} 
\vdots \\
X_p \nabla \left( I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\#
(X_{p'}) \right) \\
\vdots
\end{array} \right]$$

@@

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence could be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilizing innermost loops, or stabilizing strongly
connected components of the graph.

@@

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, $i \in
[0,1]$, then $i\in [0,2]$ after the second iteration\dots. 
After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better. We can repeat such iterations as long as the set narrows, and
we always obtain a correct over-approximation.
In practice, a single narrowing iteration is enough to recover the precision.

@@

To limit the number of such points, a method could be
to expand the control flow graph. For instance, instead of considering a graph
with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, 
the $n$ merge node
of these \emph{if-then-else} could be removed, each of them having two incoming
edges, and 
$2^n$ edges from $p_1$ to $p_n$ are added, corresponding to the different paths through
the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
several incoming edges. This graph transformation results in an exponential
blowup, because of the exponential growth of the number of edges in the graph.
\end{enumerate}

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
is aimed at computing the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, an invariant could be directly computed for the loop:

@@

Lookahead widening \cite{GopanR06} is a technique that isolates the
different loop phases during the analysis, and stabilizes each of these loop
phases before proceeding to the next. One only consider paths that are feasible
at the first iteration, and forget the others for a while. The iterations end
when the analysis of these paths has converged. After that, some
narrowing iterations can be computed. Then, paths that
become feasible are reinserted into the control flow graph, and we redo
iterations.

@@

In the classical abstract interpretation technique, narrowing iterations are
computed at the end of the analysis, when all the fixpoint computation has
converged.
The interest of this method is to apply narrowing iterations right after each
path computation. This allows to recover precision before analyzing new paths,
and then to be more precise.

@@

	Intuitively, expanding the graph comes back to consider independently the
	different paths in the control flow graph, and thus to be more precise,
	because paths that are not feasible will be ignored. Additionally, the
	number of control points with several incoming transitions will be reduced,
	so the loss of precision due to the convex hull of polyhedra may be
	minimized.

@@

A $\Phi$-function is usually inserted at points of the control flow graph that
have several incoming transitions, and has the same number of arguments as the point
has incoming transitions. If we suppose these transitions are ordered, then the
$\Phi$-function returns the value of its $i$-th argument if the point is reached
from the $i$-th transition.

@@

	\begin{center} \emph{
		``Is there a path starting on control point $p$, with numerical
		variables in $X_p$, that ends on point $q$, with variables that are not
		in $X_q$ ?''}
	\end{center}

@@

	Although deciding the satisfiability of such formula is NP-complete, there
	has been much research on decision procedures \cite{Kroening08} and there exist nowadays
	efficient programs, known as SMT-solvers, that can decide the satisfiability
	of LRA or LIA formulas. Well-known efficient SMT-solvers are
	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantic, expressed in
terms of the boolean and numerical SSA variables. For instance, if the CFG
contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ express the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a boolean variable left non-deterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, standardly defined in the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	To compute the reachability strating from point $p_i$, the formula given to
	the SMT-solver is:

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator,  some narrowing steps could be performed to regain precision
	before focusing on another path. This is the same principle as the one we
	find in \emph{Lookahead widening} \cite{GopanR06}.

@@

	When the self loop is accelerable \cite{Gon07}, instead of doing a
	widening/narrowing sequence,  acceleration directly finds
	its transitive closure.

@@

>>>>

\usepackage{verimagstudent}
\usepackage{mathcomp,amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix} 
% DM: verimagstudent charge hyperref, et hyperref doit habituellement
% DM: passer en dernier
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\Q{\mathbb{Q}}
\def\P{\mathcal{P}}
\def\ite{\textnormal{ite\ }}
\def\lfp{\textnormal{\it lfp}}
\newcommand{\widening}{\mathop{\triangledown}}

@@

\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aims at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

Static analysis aims at automatically computing properties on programs, such as
the possible values of their variables during execution. This allows to show for
instance that a program will not overflow, will not divide by zero, and to compute
loop invariants\dots
The main applications of static analysis are compilation-time optimisation, and
the proof of safety properties in critical systems, such as avionics.

@@

Satisfiability Modulo Theory (SMT) solving is a technique
for deciding the satisfiability of a logic formula containing Boolean
predicates and elements of a theory, such as linear arithmetic relations between
integers.
In this report, we present a novel refinement of the Abstract Interpretation
technique, that uses SMT-solving to guide the fixpoint iterations, so that it
temporarily focuses on a selected path in the control flow graph (the graph
expressing the semantics of the program), in order to
obtain more precise results at reasonable cost.

\subsection*{Contribution and Organisation}

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, referred as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyser, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

The report is organised as follows:
Part \ref{stateoftheartpart} introduces the state of the art in Abstract
Interpretation and more specifically in Linear Relation Analysis.
Part \ref{pathfocusingpart} describes the two techniques
\emph{Lookahead Widening} and \emph{Path Focusing}.
Part \ref{implementationpart} presents the implementation of these two
techniques into an analyser, and the experimental results we obtained.
Finally, Part \ref{future} introduces some directions to explore.

@@

\section{Introductory example}

@@

There is a relation between $Y_1$ and $Y_2$ :
\begin{eqnarray*}
Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
\end{eqnarray*}
% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
% DM: et pas de depart
We see that $Y_1$ and $Y_2$ can be computed step by step:
we start with $Y_1 = Y_2 = \emptyset$.
Then, we add iteratively new elements in the sets.
\begin{itemize}
\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
\{0\}$.
\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
\item Again, we find $Y_2 = \{0,1\}$.
\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
\item We continue the iteration until we find no more elements in the sets\dots
\end{itemize}

@@

Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method for
finding approximate solutions of
fixpoint equations. This method is used for program analysis, since analysing a
program often comes down to solving fixpoint equations, because of the loops and
recursive procedures.

@@

Abstract Interpretation proposes to represent more efficiently the
elements of this complex domain $C$ of concrete values,
by choosing an simpler domain  $A$ called \emph{abstract domain}. 
% DM: un peu lourd et peu clair

@@

Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
= \gamma \circ \Phi^\#(x)$. Abstract interpretation aims at computing
such an~$x$. It computes the stationary limit of an ascending sequence
$(x_i)_{i \geq 0}$ defined by the induction:

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$
where $\perp$ is the least element of $A$.
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, of finite height,
or more generally satisfy the \emph{ascending chain condition},
meaning that there doesn't exist
any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, otherwise, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

Many of the domains that are used for program analysis, including those of intervals and convex polyhedra, don't satisfy this ascending chain condition. To ensure convergence in this case, another approximation
is performed: a new operator is defined, called \emph{widening operator}, that
extrapolates the limit of a sequence of abstract values
\cite{CC77,CousotCousot92-4}. This \emph{widening
operator} is usually noted $\widening: A \times A \rightarrow A$, and satisfies
the following properties:

\begin{itemize}
\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
\widening x_2$. This guarantees the correctness of the result.
% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
% DM: une union abstraite juste avant
\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
defined by 
$$\system{
x'_0 & = &  x_0 \\
x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i > 0
}$$
is not strictly increasing. Then, applying the widening operator when the
sequence may increase indefinitely makes the computation converge to a fixpoint
in finite time. 
\end{itemize}

\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.
% DM: c'est gentil de me citer!

@@

$$\system{
x'_0 &=& \perp \\
x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i > 0
}$$

@@

Since the domain of convex polyhedra is of infinite height, a widening operator
is defined to ensure convergence of the technique. 
Intuitively, the polyhedron $P \widening Q$ is obtained after 
removing from the system of linear equations defining $P$ all the inequalities
that are not satisfied by $Q$ (the actual definition is somewhat more involved,
see e.g. \cite{BagnaraHRZ05SCP}).
For instance, if $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$ and $Q = 
\{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$, then the result of the widening operator
will be $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.

There have been various
work proposing a definition or a refinement of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying widening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. Most of the
cases, $n$ can be chosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\widening_T$ is defined as follows:
$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

For instance, consider the program:
\begin{C}
while (x <= 50) {
	x++;
	y++;
}
\end{C}
If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
widening (which gives the result $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
a widening with the threshold $x \leq 50$ gives 
$P \widening_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
with threshold helps to find better invariants.
\end{itemize}

@@

$$\left[ \begin{array}{c}
\vdots \\
X_p \\
\vdots
\end{array} \right] \longleftarrow \left[ \begin{array}{c} 
\vdots \\
X_p \widening \left( I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\#
(X_{p'}) \right) \\
\vdots
\end{array} \right]$$

@@

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence could be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilising innermost loops, or stabilising strongly
connected components of the graph.

@@

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, $i \in
[0,1]$, then $i\in [0,2]$ after the second iteration\dots. 
After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better. We can repeat such iterations as long as the set narrows, and
we always obtain a correct over-approximation.
In practise, a single narrowing iteration is enough to recover the precision.

@@

To limit the number of such points, a method could be
to expand the control flow graph. For instance, instead of considering a graph
with a sequence of $n$ \emph{if-then-else} between points $p_1$ and $p_n$, 
the $n$ merge node
of these \emph{if-then-else} could be removed, each of them having two incoming
edges, and 
$2^n$ edges from $p_1$ to $p_n$ are added, corresponding to the different paths through
the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
several incoming edges. This graph transformation results in an exponential
blowup, because of the exponential growth of the number of edges in the graph.
\end{enumerate}

@@

Typically, to analyse a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
is aimed at computing the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, an invariant could be directly computed for the loop:

@@

Lookahead widening \cite{GopanR06} is a technique that isolates the
different loop phases during the analysis, and stabilises each of these loop
phases before proceeding to the next. One only consider paths that are feasible
at the first iteration, and forget the others for a while. The iterations end
when the analysis of these paths has converged. After that, some
narrowing iterations can be computed. Then, paths that
become feasible are reinserted into the control flow graph, and we redo
iterations.

@@

In the classical abstract interpretation technique, narrowing iterations are
computed at the end of the analysis, when all the fixpoint computation has
converged.
The interest of this method is to apply narrowing iterations right after each
path computation. This allows to recover precision before analysing new paths,
and then to be more precise.

@@

	Intuitively, expanding the graph comes back to consider independently the
	different paths in the control flow graph, and thus to be more precise,
	because paths that are not feasible will be ignored. Additionally, the
	number of control points with several incoming transitions will be reduced,
	so the loss of precision due to the convex hull of polyhedra may be
	minimised.

@@

A $\Phi$-function is usually inserted at points of the control flow graph that
have several incoming transitions, and has the same number of arguments as the point
has incoming transitions. Assuming these transitions are ordered, the
$\Phi$-function returns the value of its $i$-th argument if the point is reached
from the $i$-th transition.

@@

	\begin{center} \emph{
		``Is there a path starting on control point $p$, with numerical
		variables in $X_p$, that ends on point $q$, with variables that are not
		in $X_q$?''}
	\end{center}

@@

	Although deciding the satisfiability of such formula is NP-complete, there
	has been much research on decision procedures \cite{Kroening08} and there exist nowadays
	efficient programs, known as SMT-solvers, that can decide the satisfiability
	of LRA or LIA formulae. Well-known efficient SMT-solvers are
	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantic, expressed in
terms of the Boolean and numerical SSA variables. For instance, if the CFG
contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ expresses the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a Boolean variable left non-deterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, a standard feature of the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	To compute the reachability starting from point $p_i$, the formula given to
	the SMT-solver is:

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator,  some narrowing steps could be performed to regain precision
	before focusing on another path. This relates to the ideas in
        \emph{Lookahead widening} \cite{GopanR06}.

@@

	When the self loop can be accelerated \cite{Gon07}, instead of doing a
	widening/narrowing sequence,  acceleration directly finds
	its transitive closure.

@@

%DM: en quoi tout ceci est de la transitive closure?
%DM: transitive closure c'est sur des relations i/o
%DM: note que tu pourrais faire des relations (summaries de boucles, etc.)
%DM: en gardant dans tes polyedres une copie des variables en tete
%DM: du bloc/fonction/truc
